---
title: "Movielens Project"
author: "George Moisescu"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

``` {r, echo = FALSE, message=FALSE, warning=FALSE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)
library(lubridate)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                           title = as.character(title),
                                           genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)

```


# **Introduction**

  This is my final project within the Data Science program and I try to make a movies recommendation algorithm which will predict the movie rating based on the userId, movieId, year of rating and movie genre. I use the movilens dataset provided by GroupLens research lab.

```{r}

head(edx)

```
# **Exploratory Data Analysis**

  As we can see the data is not in tidy format, there are timestamp variable which contain  the number of seconds since 1970-01-01 and the genres variable which contain several genre in the same observation.
  
```{r,message=FALSE, warning=FALSE}

max(str_count(edx$genres, "\\|"))
edx_tidy <- edx %>% separate(genres,c("1_genre","2_genre","3_genre","4_genre",
                                      "5_genre","6_genre","7_genre","8_genre"),"\\|")

individual_genre <- unique(c(unique(edx_tidy$`1_genre`),unique(edx_tidy$`2_genre`),
                             unique(edx_tidy$`3_genre`),unique(edx_tidy$`4_genre`),
                             unique(edx_tidy$`5_genre`),unique(edx_tidy$`6_genre`),
                             unique(edx_tidy$`7_genre`),unique(edx_tidy$`8_genre`)))
individual_genre <-individual_genre[order(individual_genre)]


individual_genre

```  
  We have 20 individual genres. I create 20 new columns and I populate them width 1 if the movie belong to the that genre or 0 if it doesn't.I extract also the year in which the rate was done from the timestamp column.
  
```{r,message=FALSE, warning=FALSE}

edx_tidy <- edx_tidy %>% mutate(Action = 0,Adventure = 0,Animation = 0,Children = 0,
                                Comedy = 0,Crime = 0,Documentary = 0,Drama = 0,
                                Fantasy = 0,Film_Noir = 0,Horror = 0,IMAX = 0,
                                Musical = 0,Mystery = 0,Romance = 0,Sci_Fi = 0,
                                Thriller = 0,War = 0, Western = 0,No_Genre = 0)

temp <- edx_tidy$`1_genre` == "Action"
edx_tidy$Action[temp] <- 1

temp <- edx_tidy$`1_genre` == "Adventure"
edx_tidy$Adventure[temp] <- 1

```

```{r, echo=FALSE,message=FALSE,warning=FALSE}

temp <- edx_tidy$`1_genre` == "Animation"
edx_tidy$Animation[temp] <- 1

temp <- edx_tidy$`1_genre` == "Children"
edx_tidy$Children[temp] <- 1

temp <- edx_tidy$`1_genre` == "Comedy"
edx_tidy$Comedy[temp] <- 1

temp <- edx_tidy$`1_genre` == "Crime"
edx_tidy$Crime[temp] <- 1

temp <- edx_tidy$`1_genre` == "Documentary"
edx_tidy$Documentary[temp] <- 1

temp <- edx_tidy$`1_genre` == "Drama"
edx_tidy$Drama[temp] <- 1

temp <- edx_tidy$`1_genre` == "Fantasy"
edx_tidy$Fantasy[temp] <- 1

temp <- edx_tidy$`1_genre` == "Film-Noir"
edx_tidy$Film_Noir[temp] <- 1

temp <- edx_tidy$`1_genre` == "Horror"
edx_tidy$Horror[temp] <- 1

temp <- edx_tidy$`1_genre` == "IMAX"
edx_tidy$IMAX[temp] <- 1

temp <- edx_tidy$`1_genre` == "Musical"
edx_tidy$Musical[temp] <- 1

temp <- edx_tidy$`1_genre` == "Mystery"
edx_tidy$Mystery[temp] <- 1

temp <- edx_tidy$`1_genre` == "Romance"
edx_tidy$Romance[temp] <- 1

temp <- edx_tidy$`1_genre` == "Sci-Fi"
edx_tidy$Sci_Fi[temp] <- 1

temp <- edx_tidy$`1_genre` == "Thriller"
edx_tidy$Thriller[temp] <- 1

temp <- edx_tidy$`1_genre` == "War"
edx_tidy$War[temp] <- 1

temp <- edx_tidy$`1_genre` == "Western"
edx_tidy$Western[temp] <- 1

temp <- edx_tidy$`1_genre` == "(no genres listed)"
edx_tidy$No_Genre[temp] <- 1




temp <- edx_tidy$`2_genre` == "Action"
edx_tidy$Action[temp] <- 1

temp <- edx_tidy$`2_genre` == "Adventure"
edx_tidy$Adventure[temp] <- 1

temp <- edx_tidy$`2_genre` == "Animation"
edx_tidy$Animation[temp] <- 1

temp <- edx_tidy$`2_genre` == "Children"
edx_tidy$Children[temp] <- 1

temp <- edx_tidy$`2_genre` == "Comedy"
edx_tidy$Comedy[temp] <- 1

temp <- edx_tidy$`2_genre` == "Crime"
edx_tidy$Crime[temp] <- 1

temp <- edx_tidy$`2_genre` == "Documentary"
edx_tidy$Documentary[temp] <- 1

temp <- edx_tidy$`2_genre` == "Drama"
edx_tidy$Drama[temp] <- 1

temp <- edx_tidy$`2_genre` == "Fantasy"
edx_tidy$Fantasy[temp] <- 1

temp <- edx_tidy$`2_genre` == "Film-Noir"
edx_tidy$Film_Noir[temp] <- 1

temp <- edx_tidy$`2_genre` == "Horror"
edx_tidy$Horror[temp] <- 1

temp <- edx_tidy$`2_genre` == "IMAX"
edx_tidy$IMAX[temp] <- 1

temp <- edx_tidy$`2_genre` == "Musical"
edx_tidy$Musical[temp] <- 1

temp <- edx_tidy$`2_genre` == "Mystery"
edx_tidy$Mystery[temp] <- 1

temp <- edx_tidy$`2_genre` == "Romance"
edx_tidy$Romance[temp] <- 1

temp <- edx_tidy$`2_genre` == "Sci-Fi"
edx_tidy$Sci_Fi[temp] <- 1

temp <- edx_tidy$`2_genre` == "Thriller"
edx_tidy$Thriller[temp] <- 1

temp <- edx_tidy$`2_genre` == "War"
edx_tidy$War[temp] <- 1

temp <- edx_tidy$`2_genre` == "Western"
edx_tidy$Western[temp] <- 1

temp <- edx_tidy$`2_genre` == "(no genres listed)"
edx_tidy$No_Genre[temp] <- 1




temp <- edx_tidy$`3_genre` == "Action"
edx_tidy$Action[temp] <- 1

temp <- edx_tidy$`3_genre` == "Adventure"
edx_tidy$Adventure[temp] <- 1

temp <- edx_tidy$`3_genre` == "Animation"
edx_tidy$Animation[temp] <- 1

temp <- edx_tidy$`3_genre` == "Children"
edx_tidy$Children[temp] <- 1

temp <- edx_tidy$`3_genre` == "Comedy"
edx_tidy$Comedy[temp] <- 1

temp <- edx_tidy$`3_genre` == "Crime"
edx_tidy$Crime[temp] <- 1

temp <- edx_tidy$`3_genre` == "Documentary"
edx_tidy$Documentary[temp] <- 1

temp <- edx_tidy$`3_genre` == "Drama"
edx_tidy$Drama[temp] <- 1

temp <- edx_tidy$`3_genre` == "Fantasy"
edx_tidy$Fantasy[temp] <- 1

temp <- edx_tidy$`3_genre` == "Film-Noir"
edx_tidy$Film_Noir[temp] <- 1

temp <- edx_tidy$`3_genre` == "Horror"
edx_tidy$Horror[temp] <- 1

temp <- edx_tidy$`3_genre` == "IMAX"
edx_tidy$IMAX[temp] <- 1

temp <- edx_tidy$`3_genre` == "Musical"
edx_tidy$Musical[temp] <- 1

temp <- edx_tidy$`3_genre` == "Mystery"
edx_tidy$Mystery[temp] <- 1

temp <- edx_tidy$`3_genre` == "Romance"
edx_tidy$Romance[temp] <- 1

temp <- edx_tidy$`3_genre` == "Sci-Fi"
edx_tidy$Sci_Fi[temp] <- 1

temp <- edx_tidy$`3_genre` == "Thriller"
edx_tidy$Thriller[temp] <- 1

temp <- edx_tidy$`3_genre` == "War"
edx_tidy$War[temp] <- 1

temp <- edx_tidy$`3_genre` == "Western"
edx_tidy$Western[temp] <- 1

temp <- edx_tidy$`3_genre` == "(no genres listed)"
edx_tidy$No_Genre[temp] <- 1



temp <- edx_tidy$`4_genre` == "Action"
edx_tidy$Action[temp] <- 1

temp <- edx_tidy$`4_genre` == "Adventure"
edx_tidy$Adventure[temp] <- 1

temp <- edx_tidy$`4_genre` == "Animation"
edx_tidy$Animation[temp] <- 1

temp <- edx_tidy$`4_genre` == "Children"
edx_tidy$Children[temp] <- 1

temp <- edx_tidy$`4_genre` == "Comedy"
edx_tidy$Comedy[temp] <- 1

temp <- edx_tidy$`4_genre` == "Crime"
edx_tidy$Crime[temp] <- 1

temp <- edx_tidy$`4_genre` == "Documentary"
edx_tidy$Documentary[temp] <- 1

temp <- edx_tidy$`4_genre` == "Drama"
edx_tidy$Drama[temp] <- 1

temp <- edx_tidy$`4_genre` == "Fantasy"
edx_tidy$Fantasy[temp] <- 1

temp <- edx_tidy$`4_genre` == "Film-Noir"
edx_tidy$Film_Noir[temp] <- 1

temp <- edx_tidy$`4_genre` == "Horror"
edx_tidy$Horror[temp] <- 1

temp <- edx_tidy$`4_genre` == "IMAX"
edx_tidy$IMAX[temp] <- 1

temp <- edx_tidy$`4_genre` == "Musical"
edx_tidy$Musical[temp] <- 1

temp <- edx_tidy$`4_genre` == "Mystery"
edx_tidy$Mystery[temp] <- 1

temp <- edx_tidy$`4_genre` == "Romance"
edx_tidy$Romance[temp] <- 1

temp <- edx_tidy$`4_genre` == "Sci-Fi"
edx_tidy$Sci_Fi[temp] <- 1

temp <- edx_tidy$`4_genre` == "Thriller"
edx_tidy$Thriller[temp] <- 1

temp <- edx_tidy$`4_genre` == "War"
edx_tidy$War[temp] <- 1

temp <- edx_tidy$`4_genre` == "Western"
edx_tidy$Western[temp] <- 1

temp <- edx_tidy$`4_genre` == "(no genres listed)"
edx_tidy$No_Genre[temp] <- 1



temp <- edx_tidy$`5_genre` == "Action"
edx_tidy$Action[temp] <- 1

temp <- edx_tidy$`5_genre` == "Adventure"
edx_tidy$Adventure[temp] <- 1

temp <- edx_tidy$`5_genre` == "Animation"
edx_tidy$Animation[temp] <- 1

temp <- edx_tidy$`5_genre` == "Children"
edx_tidy$Children[temp] <- 1

temp <- edx_tidy$`5_genre` == "Comedy"
edx_tidy$Comedy[temp] <- 1

temp <- edx_tidy$`5_genre` == "Crime"
edx_tidy$Crime[temp] <- 1

temp <- edx_tidy$`5_genre` == "Documentary"
edx_tidy$Documentary[temp] <- 1

temp <- edx_tidy$`5_genre` == "Drama"
edx_tidy$Drama[temp] <- 1

temp <- edx_tidy$`5_genre` == "Fantasy"
edx_tidy$Fantasy[temp] <- 1

temp <- edx_tidy$`5_genre` == "Film-Noir"
edx_tidy$Film_Noir[temp] <- 1

temp <- edx_tidy$`5_genre` == "Horror"
edx_tidy$Horror[temp] <- 1

temp <- edx_tidy$`5_genre` == "IMAX"
edx_tidy$IMAX[temp] <- 1

temp <- edx_tidy$`5_genre` == "Musical"
edx_tidy$Musical[temp] <- 1

temp <- edx_tidy$`5_genre` == "Mystery"
edx_tidy$Mystery[temp] <- 1

temp <- edx_tidy$`5_genre` == "Romance"
edx_tidy$Romance[temp] <- 1

temp <- edx_tidy$`5_genre` == "Sci-Fi"
edx_tidy$Sci_Fi[temp] <- 1

temp <- edx_tidy$`5_genre` == "Thriller"
edx_tidy$Thriller[temp] <- 1

temp <- edx_tidy$`5_genre` == "War"
edx_tidy$War[temp] <- 1

temp <- edx_tidy$`5_genre` == "Western"
edx_tidy$Western[temp] <- 1

temp <- edx_tidy$`5_genre` == "(no genres listed)"
edx_tidy$No_Genre[temp] <- 1



temp <- edx_tidy$`6_genre` == "Action"
edx_tidy$Action[temp] <- 1

temp <- edx_tidy$`6_genre` == "Adventure"
edx_tidy$Adventure[temp] <- 1

temp <- edx_tidy$`6_genre` == "Animation"
edx_tidy$Animation[temp] <- 1

temp <- edx_tidy$`6_genre` == "Children"
edx_tidy$Children[temp] <- 1

temp <- edx_tidy$`6_genre` == "Comedy"
edx_tidy$Comedy[temp] <- 1

temp <- edx_tidy$`6_genre` == "Crime"
edx_tidy$Crime[temp] <- 1

temp <- edx_tidy$`6_genre` == "Documentary"
edx_tidy$Documentary[temp] <- 1

temp <- edx_tidy$`6_genre` == "Drama"
edx_tidy$Drama[temp] <- 1

temp <- edx_tidy$`6_genre` == "Fantasy"
edx_tidy$Fantasy[temp] <- 1

temp <- edx_tidy$`6_genre` == "Film-Noir"
edx_tidy$Film_Noir[temp] <- 1

temp <- edx_tidy$`6_genre` == "Horror"
edx_tidy$Horror[temp] <- 1

temp <- edx_tidy$`6_genre` == "IMAX"
edx_tidy$IMAX[temp] <- 1

temp <- edx_tidy$`6_genre` == "Musical"
edx_tidy$Musical[temp] <- 1

temp <- edx_tidy$`6_genre` == "Mystery"
edx_tidy$Mystery[temp] <- 1

temp <- edx_tidy$`6_genre` == "Romance"
edx_tidy$Romance[temp] <- 1

temp <- edx_tidy$`6_genre` == "Sci-Fi"
edx_tidy$Sci_Fi[temp] <- 1

temp <- edx_tidy$`6_genre` == "Thriller"
edx_tidy$Thriller[temp] <- 1

temp <- edx_tidy$`6_genre` == "War"
edx_tidy$War[temp] <- 1

temp <- edx_tidy$`6_genre` == "Western"
edx_tidy$Western[temp] <- 1

temp <- edx_tidy$`6_genre` == "(no genres listed)"
edx_tidy$No_Genre[temp] <- 1


temp <- edx_tidy$`7_genre` == "Action"
edx_tidy$Action[temp] <- 1

temp <- edx_tidy$`7_genre` == "Adventure"
edx_tidy$Adventure[temp] <- 1

temp <- edx_tidy$`7_genre` == "Animation"
edx_tidy$Animation[temp] <- 1

temp <- edx_tidy$`7_genre` == "Children"
edx_tidy$Children[temp] <- 1

temp <- edx_tidy$`7_genre` == "Comedy"
edx_tidy$Comedy[temp] <- 1

temp <- edx_tidy$`7_genre` == "Crime"
edx_tidy$Crime[temp] <- 1

temp <- edx_tidy$`7_genre` == "Documentary"
edx_tidy$Documentary[temp] <- 1

temp <- edx_tidy$`7_genre` == "Drama"
edx_tidy$Drama[temp] <- 1

temp <- edx_tidy$`7_genre` == "Fantasy"
edx_tidy$Fantasy[temp] <- 1

temp <- edx_tidy$`7_genre` == "Film-Noir"
edx_tidy$Film_Noir[temp] <- 1

temp <- edx_tidy$`7_genre` == "Horror"
edx_tidy$Horror[temp] <- 1

temp <- edx_tidy$`7_genre` == "IMAX"
edx_tidy$IMAX[temp] <- 1

temp <- edx_tidy$`7_genre` == "Musical"
edx_tidy$Musical[temp] <- 1

temp <- edx_tidy$`7_genre` == "Mystery"
edx_tidy$Mystery[temp] <- 1

temp <- edx_tidy$`7_genre` == "Romance"
edx_tidy$Romance[temp] <- 1

temp <- edx_tidy$`7_genre` == "Sci-Fi"
edx_tidy$Sci_Fi[temp] <- 1

temp <- edx_tidy$`7_genre` == "Thriller"
edx_tidy$Thriller[temp] <- 1

temp <- edx_tidy$`7_genre` == "War"
edx_tidy$War[temp] <- 1

temp <- edx_tidy$`7_genre` == "Western"
edx_tidy$Western[temp] <- 1

temp <- edx_tidy$`7_genre` == "(no genres listed)"
edx_tidy$No_Genre[temp] <- 1



temp <- edx_tidy$`8_genre` == "Action"
edx_tidy$Action[temp] <- 1

temp <- edx_tidy$`8_genre` == "Adventure"
edx_tidy$Adventure[temp] <- 1

temp <- edx_tidy$`8_genre` == "Animation"
edx_tidy$Animation[temp] <- 1

temp <- edx_tidy$`8_genre` == "Children"
edx_tidy$Children[temp] <- 1

temp <- edx_tidy$`8_genre` == "Comedy"
edx_tidy$Comedy[temp] <- 1

temp <- edx_tidy$`8_genre` == "Crime"
edx_tidy$Crime[temp] <- 1

temp <- edx_tidy$`8_genre` == "Documentary"
edx_tidy$Documentary[temp] <- 1

temp <- edx_tidy$`8_genre` == "Drama"
edx_tidy$Drama[temp] <- 1

temp <- edx_tidy$`8_genre` == "Fantasy"
edx_tidy$Fantasy[temp] <- 1

temp <- edx_tidy$`8_genre` == "Film-Noir"
edx_tidy$Film_Noir[temp] <- 1

temp <- edx_tidy$`8_genre` == "Horror"
edx_tidy$Horror[temp] <- 1

temp <- edx_tidy$`8_genre` == "IMAX"
edx_tidy$IMAX[temp] <- 1

temp <- edx_tidy$`8_genre` == "Musical"
edx_tidy$Musical[temp] <- 1

temp <- edx_tidy$`8_genre` == "Mystery"
edx_tidy$Mystery[temp] <- 1

temp <- edx_tidy$`8_genre` == "Romance"
edx_tidy$Romance[temp] <- 1

temp <- edx_tidy$`8_genre` == "Sci-Fi"
edx_tidy$Sci_Fi[temp] <- 1

temp <- edx_tidy$`8_genre` == "Thriller"
edx_tidy$Thriller[temp] <- 1

```
---------------- etc.

```{r,message=FALSE,warning=FALSE}
temp <- edx_tidy$`8_genre` == "War"
edx_tidy$War[temp] <- 1

temp <- edx_tidy$`8_genre` == "Western"
edx_tidy$Western[temp] <- 1

temp <- edx_tidy$`8_genre` == "(no genres listed)"
edx_tidy$No_Genre[temp] <- 1
```

```{r,message=FALSE,warning=FALSE}

edx_tidy <- edx_tidy[,-c(6:13)]
edx_tidy <- edx_tidy%>% mutate(rated_year = year(as_datetime(edx_tidy$timestamp)))
edx_tidy <- edx_tidy[,-4]
edx_tidy <- edx_tidy[,c(1:3,25,4:24)]
head(edx_tidy)

```
## **a. Genres Analysis**

```{r} 
genre_sum <- rep(0,20)
genre_rating <- rep(0,20)

genre_sum[1] <- sum(edx_tidy$Action)
  filtered_data <- edx_tidy %>% filter(Action == 1)
  genre_rating[1] <- round(mean(filtered_data$rating),digits = 2)

  genre_sum[2] <- sum(edx_tidy$Adventure)
  filtered_data <- edx_tidy %>% filter(Adventure == 1)
  genre_rating[2] <- round(mean(filtered_data$rating),digits = 2)
```
--------------etc.

```{r,echo=FALSE,message=FALSE,warning=FALSE}
 genre_sum[3] <- sum(edx_tidy$Animation)
  filtered_data <- edx_tidy %>% filter(Animation == 1)
  genre_rating[3] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[4] <- sum(edx_tidy$Children)
  filtered_data <- edx_tidy %>% filter(Children == 1)
  genre_rating[4] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[5] <- sum(edx_tidy$Comedy)
  filtered_data <- edx_tidy %>% filter(Comedy == 1)
  genre_rating[5] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[6] <- sum(edx_tidy$Crime)
  filtered_data <- edx_tidy %>% filter(Crime == 1)
  genre_rating[6] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[7] <- sum(edx_tidy$Documentary)
  filtered_data <- edx_tidy %>% filter(Documentary == 1)
  genre_rating[7] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[8] <- sum(edx_tidy$Drama)
  filtered_data <- edx_tidy %>% filter(Drama == 1)
  genre_rating[8] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[9] <- sum(edx_tidy$Fantasy)
  filtered_data <- edx_tidy %>% filter(Fantasy == 1)
  genre_rating[9] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[10] <- sum(edx_tidy$Film_Noir)
  filtered_data <- edx_tidy %>% filter(Film_Noir == 1)
  genre_rating[10] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[11] <- sum(edx_tidy$Horror)
  filtered_data <- edx_tidy %>% filter(Horror == 1)
  genre_rating[11] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[12] <- sum(edx_tidy$IMAX)
  filtered_data <- edx_tidy %>% filter(IMAX == 1)
  genre_rating[12] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[13] <- sum(edx_tidy$Musical)
  filtered_data <- edx_tidy %>% filter(Musical == 1)
  genre_rating[13] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[14] <- sum(edx_tidy$Mystery)
  filtered_data <- edx_tidy %>% filter(Mystery == 1)
  genre_rating[14] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[15] <- sum(edx_tidy$Romance)
  filtered_data <- edx_tidy %>% filter(Romance == 1)
  genre_rating[15] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[16] <- sum(edx_tidy$Sci_Fi)
  filtered_data <- edx_tidy %>% filter(Sci_Fi == 1)
  genre_rating[16] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[17] <- sum(edx_tidy$Thriller)
  filtered_data <- edx_tidy %>% filter(Thriller == 1)
  genre_rating[17] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[18] <- sum(edx_tidy$War)
  filtered_data <- edx_tidy %>% filter(War == 1)
  genre_rating[18] <- round(mean(filtered_data$rating),digits = 2)
  
  genre_sum[19] <- sum(edx_tidy$Western)
  filtered_data <- edx_tidy %>% filter(Western == 1)
  genre_rating[19] <- round(mean(filtered_data$rating),digits = 2)
```
```{r,message=FALSE,warning=FALSE}

  genre_sum[20] <- sum(edx_tidy$No_Genre)
  filtered_data <- edx_tidy %>% filter(No_Genre == 1)
  genre_rating[20] <- round(mean(filtered_data$rating),digits = 2)
  
```
```{r,message=FALSE,warning=FALSE}
genre_data <- data.frame(genre = colnames(edx_tidy[,-c(1:5)]),no_of_ratings = 
                                  genre_sum,rating_mean = genre_rating)
genre_data <- genre_data[order(genre_data$no_of_ratings,decreasing = TRUE),]
genre_data
genre_data %>% ggplot(aes(genre,no_of_ratings)) + geom_point(size = 3, color = "blue") +
  theme(axis.text.x = element_text(angle = 90, face = "bold",vjust = 0.5,hjust = 1)) +
  ggtitle("Variation of number of ratings with genre category")

```
 
 We can see that there are some genres which are rated more than others. The Drama, Comedy and Action are the most three rated genres. The average rating varies between 3.27 for Horror movies and 4.01 for Film_Noir movies. The Film_Noir movies are on the 17th place in the number of rating list. There is a small variation regarding average rating between the movie genre.
 
## **b. Rated Years Analysis**
 
 ```{r,message=FALSE, warning=FALSE}
 
 year_data <- edx_tidy %>% group_by(rated_year) %>% summarize(no_of_ratings = n(),
                                                    average_rating = mean(rating))
year_data[order(year_data$no_of_ratings,decreasing = TRUE),]
year_data %>% ggplot(aes(factor(rated_year),no_of_ratings)) + geom_point(size = 3, 
                                                              color = "blue") +
  theme(axis.text.x = element_text(angle = 90, face = "bold",vjust = 0.5,hjust = 1)) +
  xlab("rated_year") +
  ggtitle("Variation of number of ratings with year of ratings")
  
 ```
 
  With this summary we can see that years 2000 and 2005 has the most number of ratings. The average rating varies between 3.42 for 2004 year and 4 for 1995.

## **c. Movies Analysis**
 
```{r,message=FALSE, warning=FALSE}


movieId_data <- edx_tidy %>% group_by(movieId) %>% summarize(no_of_ratings = n(),
                                                  average_rating = mean(rating))

movieId_data <- left_join(movieId_data,unique(edx_tidy %>% select(movieId,title)),
                                                                  by = "movieId")


movieId_data <- movieId_data[order(movieId_data$no_of_ratings,decreasing = TRUE),]

head(movieId_data)
```
  There are movies that are more rated than others. Pulp Fiction, Forrest Gump and Silence of The Lambs are the first three rated movies.
  
```{r,message=FALSE, warning=FALSE}
movieId_data %>% ggplot(aes(no_of_ratings)) +
  geom_histogram(bins = 30) + 
  scale_x_log10() +
  ggtitle("The distribution of number of ratings for all movies")
```
```{r,message=FALSE, warning=FALSE}
movieId_data %>% ggplot(aes(no_of_ratings,average_rating)) +
  geom_point() +
  scale_x_log10() +
  geom_smooth(method='lm') +
  ggtitle("Variation of average rating with number of ratings of movies")

```
  We can see from this plot there is a positive influence of numbers of ratings of movies on average rating of movies.The slope of linear regression line is very small.

## **d. Users Analysis**

```{r,message=FALSE, warning=FALSE}
userId_data <- edx_tidy %>% group_by(userId) %>% summarize(no_of_ratings = n()
                                                ,average_rating = mean(rating))
userId_data <- userId_data[order(userId_data$no_of_ratings,decreasing = TRUE),]
head(userId_data)
```

There are users who give more ratings than others

```{r,message=FALSE, warning=FALSE}
userId_data %>% ggplot(aes(no_of_ratings)) +
  geom_histogram(bins = 30) +
  scale_x_log10() +
  ggtitle("Distribution of number of ratings for all users")

userId_data %>% ggplot(aes(no_of_ratings,average_rating)) +
  geom_point() +
  scale_x_log10() +
  geom_smooth(method = 'lm') +
  ggtitle("Variation of average rating with number of ratings given by users")

```


# Modeling Approach

  For this model i use the regularisation based approach. We saw from the exploratory data analysis that there are movies who a rated very few times and there are users who give very few ratings. So I put a penalty term in the  formula for calculating the biases.
  
  Accuracy of the model will be evaluated using the residual mean squared error (RMSE):
  
**_RMSE = sqrt(mean((data_rating - predicted_rating)^2))_**

* data rating = rating for movie i by user u,
* predicted rating = sum of average rating and movies biases,users biases, year biases and genre biases.


I start by predicting the same rating for all movies and all users.The value who minimizes the RMSE is the average value of all ratings.
  
```{r,message=FALSE, warning=FALSE}


validation_tidy <- validation %>% separate(genres,c("1_genre","2_genre",
                                                    "3_genre","4_genre",
                                                    "5_genre","6_genre",
                                                    "7_genre","8_genre"),
                                                    "\\|")

individual_genre <- unique(c(unique(validation_tidy$`1_genre`),
                             unique(validation_tidy$`2_genre`),
                             unique(validation_tidy$`3_genre`),
                             unique(validation_tidy$`4_genre`),
                             unique(validation_tidy$`5_genre`),
                             unique(validation_tidy$`6_genre`),
                             unique(validation_tidy$`7_genre`),
                             unique(validation_tidy$`8_genre`)))
individual_genre <-individual_genre[order(individual_genre)]

validation_tidy <- validation_tidy %>% mutate(Action = 0,Adventure = 0,
                                              Animation = 0,Children = 0,
                                              Comedy = 0,Crime = 0,
                                              Documentary = 0,Drama = 0,
                                              Fantasy = 0,Film_Noir = 0,
                                              Horror = 0,IMAX = 0,Musical = 0,
                                              Mystery = 0,Romance = 0,
                                              Sci_Fi = 0,Thriller = 0,
                                              War = 0, Western = 0,
                                              No_Genre = 0)


temp <- validation_tidy$`1_genre` == "Action"
validation_tidy$Action[temp] <- 1

temp <- validation_tidy$`1_genre` == "Adventure"
validation_tidy$Adventure[temp] <- 1

temp <- validation_tidy$`1_genre` == "Animation"
validation_tidy$Animation[temp] <- 1
```
------------------etc.

```{r,echo=FALSE,message=FALSE,warning=FALSE}
temp <- validation_tidy$`1_genre` == "Children"
validation_tidy$Children[temp] <- 1

temp <- validation_tidy$`1_genre` == "Comedy"
validation_tidy$Comedy[temp] <- 1

temp <- validation_tidy$`1_genre` == "Crime"
validation_tidy$Crime[temp] <- 1

temp <- validation_tidy$`1_genre` == "Documentary"
validation_tidy$Documentary[temp] <- 1

temp <- validation_tidy$`1_genre` == "Drama"
validation_tidy$Drama[temp] <- 1

temp <- validation_tidy$`1_genre` == "Fantasy"
validation_tidy$Fantasy[temp] <- 1

temp <- validation_tidy$`1_genre` == "Film-Noir"
validation_tidy$Film_Noir[temp] <- 1

temp <- validation_tidy$`1_genre` == "Horror"
validation_tidy$Horror[temp] <- 1

temp <- validation_tidy$`1_genre` == "IMAX"
validation_tidy$IMAX[temp] <- 1

temp <- validation_tidy$`1_genre` == "Musical"
validation_tidy$Musical[temp] <- 1

temp <- validation_tidy$`1_genre` == "Mystery"
validation_tidy$Mystery[temp] <- 1

temp <- validation_tidy$`1_genre` == "Romance"
validation_tidy$Romance[temp] <- 1

temp <- validation_tidy$`1_genre` == "Sci-Fi"
validation_tidy$Sci_Fi[temp] <- 1

temp <- validation_tidy$`1_genre` == "Thriller"
validation_tidy$Thriller[temp] <- 1

temp <- validation_tidy$`1_genre` == "War"
validation_tidy$War[temp] <- 1

temp <- validation_tidy$`1_genre` == "Western"
validation_tidy$Western[temp] <- 1

temp <- validation_tidy$`1_genre` == "(no genres listed)"
validation_tidy$No_Genre[temp] <- 1




temp <- validation_tidy$`2_genre` == "Action"
validation_tidy$Action[temp] <- 1

temp <- validation_tidy$`2_genre` == "Adventure"
validation_tidy$Adventure[temp] <- 1

temp <- validation_tidy$`2_genre` == "Animation"
validation_tidy$Animation[temp] <- 1

temp <- validation_tidy$`2_genre` == "Children"
validation_tidy$Children[temp] <- 1

temp <- validation_tidy$`2_genre` == "Comedy"
validation_tidy$Comedy[temp] <- 1

temp <- validation_tidy$`2_genre` == "Crime"
validation_tidy$Crime[temp] <- 1

temp <- validation_tidy$`2_genre` == "Documentary"
validation_tidy$Documentary[temp] <- 1

temp <- validation_tidy$`2_genre` == "Drama"
validation_tidy$Drama[temp] <- 1

temp <- validation_tidy$`2_genre` == "Fantasy"
validation_tidy$Fantasy[temp] <- 1

temp <- validation_tidy$`2_genre` == "Film-Noir"
validation_tidy$Film_Noir[temp] <- 1

temp <- validation_tidy$`2_genre` == "Horror"
validation_tidy$Horror[temp] <- 1

temp <- validation_tidy$`2_genre` == "IMAX"
validation_tidy$IMAX[temp] <- 1

temp <- validation_tidy$`2_genre` == "Musical"
validation_tidy$Musical[temp] <- 1

temp <- validation_tidy$`2_genre` == "Mystery"
validation_tidy$Mystery[temp] <- 1

temp <- validation_tidy$`2_genre` == "Romance"
validation_tidy$Romance[temp] <- 1

temp <- validation_tidy$`2_genre` == "Sci-Fi"
validation_tidy$Sci_Fi[temp] <- 1

temp <- validation_tidy$`2_genre` == "Thriller"
validation_tidy$Thriller[temp] <- 1

temp <- validation_tidy$`2_genre` == "War"
validation_tidy$War[temp] <- 1

temp <- validation_tidy$`2_genre` == "Western"
validation_tidy$Western[temp] <- 1

temp <- validation_tidy$`2_genre` == "(no genres listed)"
validation_tidy$No_Genre[temp] <- 1




temp <- validation_tidy$`3_genre` == "Action"
validation_tidy$Action[temp] <- 1

temp <- validation_tidy$`3_genre` == "Adventure"
validation_tidy$Adventure[temp] <- 1

temp <- validation_tidy$`3_genre` == "Animation"
validation_tidy$Animation[temp] <- 1

temp <- validation_tidy$`3_genre` == "Children"
validation_tidy$Children[temp] <- 1

temp <- validation_tidy$`3_genre` == "Comedy"
validation_tidy$Comedy[temp] <- 1

temp <- validation_tidy$`3_genre` == "Crime"
validation_tidy$Crime[temp] <- 1

temp <- validation_tidy$`3_genre` == "Documentary"
validation_tidy$Documentary[temp] <- 1

temp <- validation_tidy$`3_genre` == "Drama"
validation_tidy$Drama[temp] <- 1

temp <- validation_tidy$`3_genre` == "Fantasy"
validation_tidy$Fantasy[temp] <- 1

temp <- validation_tidy$`3_genre` == "Film-Noir"
validation_tidy$Film_Noir[temp] <- 1

temp <- validation_tidy$`3_genre` == "Horror"
validation_tidy$Horror[temp] <- 1

temp <- validation_tidy$`3_genre` == "IMAX"
validation_tidy$IMAX[temp] <- 1

temp <- validation_tidy$`3_genre` == "Musical"
validation_tidy$Musical[temp] <- 1

temp <- validation_tidy$`3_genre` == "Mystery"
validation_tidy$Mystery[temp] <- 1

temp <- validation_tidy$`3_genre` == "Romance"
validation_tidy$Romance[temp] <- 1

temp <- validation_tidy$`3_genre` == "Sci-Fi"
validation_tidy$Sci_Fi[temp] <- 1

temp <- validation_tidy$`3_genre` == "Thriller"
validation_tidy$Thriller[temp] <- 1

temp <- validation_tidy$`3_genre` == "War"
validation_tidy$War[temp] <- 1

temp <- validation_tidy$`3_genre` == "Western"
validation_tidy$Western[temp] <- 1

temp <- validation_tidy$`3_genre` == "(no genres listed)"
validation_tidy$No_Genre[temp] <- 1



temp <- validation_tidy$`4_genre` == "Action"
validation_tidy$Action[temp] <- 1

temp <- validation_tidy$`4_genre` == "Adventure"
validation_tidy$Adventure[temp] <- 1

temp <- validation_tidy$`4_genre` == "Animation"
validation_tidy$Animation[temp] <- 1

temp <- validation_tidy$`4_genre` == "Children"
validation_tidy$Children[temp] <- 1

temp <- validation_tidy$`4_genre` == "Comedy"
validation_tidy$Comedy[temp] <- 1

temp <- validation_tidy$`4_genre` == "Crime"
validation_tidy$Crime[temp] <- 1

temp <- validation_tidy$`4_genre` == "Documentary"
validation_tidy$Documentary[temp] <- 1

temp <- validation_tidy$`4_genre` == "Drama"
validation_tidy$Drama[temp] <- 1

temp <- validation_tidy$`4_genre` == "Fantasy"
validation_tidy$Fantasy[temp] <- 1

temp <- validation_tidy$`4_genre` == "Film-Noir"
validation_tidy$Film_Noir[temp] <- 1

temp <- validation_tidy$`4_genre` == "Horror"
validation_tidy$Horror[temp] <- 1

temp <- validation_tidy$`4_genre` == "IMAX"
validation_tidy$IMAX[temp] <- 1

temp <- validation_tidy$`4_genre` == "Musical"
validation_tidy$Musical[temp] <- 1

temp <- validation_tidy$`4_genre` == "Mystery"
validation_tidy$Mystery[temp] <- 1

temp <- validation_tidy$`4_genre` == "Romance"
validation_tidy$Romance[temp] <- 1

temp <- validation_tidy$`4_genre` == "Sci-Fi"
validation_tidy$Sci_Fi[temp] <- 1

temp <- validation_tidy$`4_genre` == "Thriller"
validation_tidy$Thriller[temp] <- 1

temp <- validation_tidy$`4_genre` == "War"
validation_tidy$War[temp] <- 1

temp <- validation_tidy$`4_genre` == "Western"
validation_tidy$Western[temp] <- 1

temp <- validation_tidy$`4_genre` == "(no genres listed)"
validation_tidy$No_Genre[temp] <- 1



temp <- validation_tidy$`5_genre` == "Action"
validation_tidy$Action[temp] <- 1

temp <- validation_tidy$`5_genre` == "Adventure"
validation_tidy$Adventure[temp] <- 1

temp <- validation_tidy$`5_genre` == "Animation"
validation_tidy$Animation[temp] <- 1

temp <- validation_tidy$`5_genre` == "Children"
validation_tidy$Children[temp] <- 1

temp <- validation_tidy$`5_genre` == "Comedy"
validation_tidy$Comedy[temp] <- 1

temp <- validation_tidy$`5_genre` == "Crime"
validation_tidy$Crime[temp] <- 1

temp <- validation_tidy$`5_genre` == "Documentary"
validation_tidy$Documentary[temp] <- 1

temp <- validation_tidy$`5_genre` == "Drama"
validation_tidy$Drama[temp] <- 1

temp <- validation_tidy$`5_genre` == "Fantasy"
validation_tidy$Fantasy[temp] <- 1

temp <- validation_tidy$`5_genre` == "Film-Noir"
validation_tidy$Film_Noir[temp] <- 1

temp <- validation_tidy$`5_genre` == "Horror"
validation_tidy$Horror[temp] <- 1

temp <- validation_tidy$`5_genre` == "IMAX"
validation_tidy$IMAX[temp] <- 1

temp <- validation_tidy$`5_genre` == "Musical"
validation_tidy$Musical[temp] <- 1

temp <- validation_tidy$`5_genre` == "Mystery"
validation_tidy$Mystery[temp] <- 1

temp <- validation_tidy$`5_genre` == "Romance"
validation_tidy$Romance[temp] <- 1

temp <- validation_tidy$`5_genre` == "Sci-Fi"
validation_tidy$Sci_Fi[temp] <- 1

temp <- validation_tidy$`5_genre` == "Thriller"
validation_tidy$Thriller[temp] <- 1

temp <- validation_tidy$`5_genre` == "War"
validation_tidy$War[temp] <- 1

temp <- validation_tidy$`5_genre` == "Western"
validation_tidy$Western[temp] <- 1

temp <- validation_tidy$`5_genre` == "(no genres listed)"
validation_tidy$No_Genre[temp] <- 1



temp <- validation_tidy$`6_genre` == "Action"
validation_tidy$Action[temp] <- 1

temp <- validation_tidy$`6_genre` == "Adventure"
validation_tidy$Adventure[temp] <- 1

temp <- validation_tidy$`6_genre` == "Animation"
validation_tidy$Animation[temp] <- 1

temp <- validation_tidy$`6_genre` == "Children"
validation_tidy$Children[temp] <- 1

temp <- validation_tidy$`6_genre` == "Comedy"
validation_tidy$Comedy[temp] <- 1

temp <- validation_tidy$`6_genre` == "Crime"
validation_tidy$Crime[temp] <- 1

temp <- validation_tidy$`6_genre` == "Documentary"
validation_tidy$Documentary[temp] <- 1

temp <- validation_tidy$`6_genre` == "Drama"
validation_tidy$Drama[temp] <- 1

temp <- validation_tidy$`6_genre` == "Fantasy"
validation_tidy$Fantasy[temp] <- 1

temp <- validation_tidy$`6_genre` == "Film-Noir"
validation_tidy$Film_Noir[temp] <- 1

temp <- validation_tidy$`6_genre` == "Horror"
validation_tidy$Horror[temp] <- 1

temp <- validation_tidy$`6_genre` == "IMAX"
validation_tidy$IMAX[temp] <- 1

temp <- validation_tidy$`6_genre` == "Musical"
validation_tidy$Musical[temp] <- 1

temp <- validation_tidy$`6_genre` == "Mystery"
validation_tidy$Mystery[temp] <- 1

temp <- validation_tidy$`6_genre` == "Romance"
validation_tidy$Romance[temp] <- 1

temp <- validation_tidy$`6_genre` == "Sci-Fi"
validation_tidy$Sci_Fi[temp] <- 1

temp <- validation_tidy$`6_genre` == "Thriller"
validation_tidy$Thriller[temp] <- 1

temp <- validation_tidy$`6_genre` == "War"
validation_tidy$War[temp] <- 1

temp <- validation_tidy$`6_genre` == "Western"
validation_tidy$Western[temp] <- 1

temp <- validation_tidy$`6_genre` == "(no genres listed)"
validation_tidy$No_Genre[temp] <- 1


temp <- validation_tidy$`7_genre` == "Action"
validation_tidy$Action[temp] <- 1

temp <- validation_tidy$`7_genre` == "Adventure"
validation_tidy$Adventure[temp] <- 1

temp <- validation_tidy$`7_genre` == "Animation"
validation_tidy$Animation[temp] <- 1

temp <- validation_tidy$`7_genre` == "Children"
validation_tidy$Children[temp] <- 1

temp <- validation_tidy$`7_genre` == "Comedy"
validation_tidy$Comedy[temp] <- 1

temp <- validation_tidy$`7_genre` == "Crime"
validation_tidy$Crime[temp] <- 1

temp <- validation_tidy$`7_genre` == "Documentary"
validation_tidy$Documentary[temp] <- 1

temp <- validation_tidy$`7_genre` == "Drama"
validation_tidy$Drama[temp] <- 1

temp <- validation_tidy$`7_genre` == "Fantasy"
validation_tidy$Fantasy[temp] <- 1

temp <- validation_tidy$`7_genre` == "Film-Noir"
validation_tidy$Film_Noir[temp] <- 1

temp <- validation_tidy$`7_genre` == "Horror"
validation_tidy$Horror[temp] <- 1

temp <- validation_tidy$`7_genre` == "IMAX"
validation_tidy$IMAX[temp] <- 1

temp <- validation_tidy$`7_genre` == "Musical"
validation_tidy$Musical[temp] <- 1

temp <- validation_tidy$`7_genre` == "Mystery"
validation_tidy$Mystery[temp] <- 1

temp <- validation_tidy$`7_genre` == "Romance"
validation_tidy$Romance[temp] <- 1

temp <- validation_tidy$`7_genre` == "Sci-Fi"
validation_tidy$Sci_Fi[temp] <- 1

temp <- validation_tidy$`7_genre` == "Thriller"
validation_tidy$Thriller[temp] <- 1

temp <- validation_tidy$`7_genre` == "War"
validation_tidy$War[temp] <- 1

temp <- validation_tidy$`7_genre` == "Western"
validation_tidy$Western[temp] <- 1

temp <- validation_tidy$`7_genre` == "(no genres listed)"
validation_tidy$No_Genre[temp] <- 1



temp <- validation_tidy$`8_genre` == "Action"
validation_tidy$Action[temp] <- 1

temp <- validation_tidy$`8_genre` == "Adventure"
validation_tidy$Adventure[temp] <- 1

temp <- validation_tidy$`8_genre` == "Animation"
validation_tidy$Animation[temp] <- 1

temp <- validation_tidy$`8_genre` == "Children"
validation_tidy$Children[temp] <- 1

temp <- validation_tidy$`8_genre` == "Comedy"
validation_tidy$Comedy[temp] <- 1

temp <- validation_tidy$`8_genre` == "Crime"
validation_tidy$Crime[temp] <- 1

temp <- validation_tidy$`8_genre` == "Documentary"
validation_tidy$Documentary[temp] <- 1

temp <- validation_tidy$`8_genre` == "Drama"
validation_tidy$Drama[temp] <- 1

temp <- validation_tidy$`8_genre` == "Fantasy"
validation_tidy$Fantasy[temp] <- 1

temp <- validation_tidy$`8_genre` == "Film-Noir"
validation_tidy$Film_Noir[temp] <- 1

temp <- validation_tidy$`8_genre` == "Horror"
validation_tidy$Horror[temp] <- 1

temp <- validation_tidy$`8_genre` == "IMAX"
validation_tidy$IMAX[temp] <- 1

temp <- validation_tidy$`8_genre` == "Musical"
validation_tidy$Musical[temp] <- 1

temp <- validation_tidy$`8_genre` == "Mystery"
validation_tidy$Mystery[temp] <- 1

temp <- validation_tidy$`8_genre` == "Romance"
validation_tidy$Romance[temp] <- 1

temp <- validation_tidy$`8_genre` == "Sci-Fi"
validation_tidy$Sci_Fi[temp] <- 1

temp <- validation_tidy$`8_genre` == "Thriller"
validation_tidy$Thriller[temp] <- 1
```

```{r,message=FALSE,warning=FALSE}
temp <- validation_tidy$`8_genre` == "War"
validation_tidy$War[temp] <- 1

temp <- validation_tidy$`8_genre` == "Western"
validation_tidy$Western[temp] <- 1

temp <- validation_tidy$`8_genre` == "(no genres listed)"
validation_tidy$No_Genre[temp] <- 1


validation_tidy <- validation_tidy[,-c(6:13)]
validation_tidy <- validation_tidy%>% mutate(rated_year = 
                    year(as_datetime(validation_tidy$timestamp)))
validation_tidy <- validation_tidy[,-4]
validation_tidy <- validation_tidy[,c(1:3,25,4:24)]

```

```{r} 


avg_rating_edx <- mean(edx_tidy$rating)


avg_rating_edx


```

## a.Movie effect

Calculating the movie bias b_i:

```{r}
movie_effect <- edx_tidy %>% group_by(movieId) %>% 
  summarize(b_i = sum(rating-avg_rating_edx)/(n()+5))
head(movie_effect)

edx_tidy <- edx_tidy %>% left_join(movie_effect, by = 'movieId')
head(edx_tidy[,c(1:4,26)])
```
Populating the validation set with the movies bias b_i:

```{r}

validation_tidy <- left_join(validation_tidy,movie_effect, by = 'movieId') 
head(validation_tidy[,c(1:4,26)])

```
## b.User effect

Calculating the user bias b_u:

```{r}
user_effect <- edx_tidy %>% group_by(userId) %>% 
  summarize(b_u = sum(rating - b_i - avg_rating_edx)/(n()+5))
head(user_effect)

edx_tidy <- edx_tidy %>% left_join(user_effect, by = 'userId')
head(edx_tidy[,c(1:4,26,27)])
```
Populating the validation set with the users bias b_u:

```{r}

validation_tidy <- left_join(validation_tidy,user_effect,by = 'userId')
head(validation_tidy[,c(1:4,26,27)])

```
## c. Year effect

Calculating the year effect b_y:

```{r}
year_effect <- edx_tidy %>% group_by(rated_year) %>% 
  summarize(b_y = sum(rating - b_i - b_u - avg_rating_edx)/(n()+5))
head(year_effect)

edx_tidy <- edx_tidy %>% left_join(year_effect, by = 'rated_year')
head(edx_tidy[,c(1:4,26,27,28)])
```
Populating the validation set with the years bias b_i:

```{r}

validation_tidy <- left_join(validation_tidy,year_effect,by = 'rated_year')
head(validation_tidy[,c(1:4,26:28)])
```
## d. Genre effect

There are 20 individual genres. The genre bias is the sum of each individual genre.

```{r}
edx_tidy <- edx_tidy %>% mutate(b_genre = 0)
validation_tidy <- validation_tidy %>% mutate(b_genre = 0)


filtered_data <- edx_tidy %>% filter(Action == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = ifelse(Action == 1,
                                                 genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = ifelse(Action == 1,
                                            genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Adventure == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u - 
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = ifelse(Adventure == 1,
                                                 genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = ifelse(Adventure == 1,
                                              genre_bias_value,0) + b_genre)
```
---------------etc.



```{r,echo=FALSE,message=FALSE,warning=FALSE}
filtered_data <- edx_tidy %>% filter(Animation == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = ifelse(Animation == 1,genre_bias_value,0)+
                                  b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                          ifelse(Animation == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Children == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                            ifelse(Children == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                          ifelse(Children == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Comedy == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                              ifelse(Comedy == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                              ifelse(Comedy == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Crime == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                                  ifelse(Crime == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                              ifelse(Crime == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Documentary == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                          ifelse(Documentary == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                        ifelse(Documentary == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Drama == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                                  ifelse(Drama == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                              ifelse(Drama == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Fantasy == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                              ifelse(Fantasy == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre =
                            ifelse(Fantasy == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Film_Noir == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                              ifelse(Film_Noir == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                          ifelse(Film_Noir == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Horror == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                  filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                            ifelse(Horror == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                              ifelse(Horror == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(IMAX == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u - 
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                                  ifelse(IMAX == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                                ifelse(IMAX == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Musical == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                            ifelse(Musical == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                            ifelse(Musical == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Mystery == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                              ifelse(Mystery == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                            ifelse(Mystery == 1,genre_bias_value,0) + b_genre)



filtered_data <- edx_tidy %>% filter(Romance == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u - 
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                            ifelse(Romance == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                            ifelse(Romance == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Sci_Fi == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                              ifelse(Sci_Fi == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre =
                              ifelse(Sci_Fi == 1,genre_bias_value,0) + b_genre)



filtered_data <- edx_tidy %>% filter(Thriller == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                  filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                            ifelse(Thriller == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                            ifelse(Thriller == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(War == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                            ifelse(War == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                                  ifelse(War == 1,genre_bias_value,0) + b_genre)


filtered_data <- edx_tidy %>% filter(Western == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                                ifelse(Western == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre = 
                            ifelse(Western == 1,genre_bias_value,0) + b_genre)
```
```{r,message=FALSE,warning=FALSE}

filtered_data <- edx_tidy %>% filter(No_Genre == 1)

effects_value <- filtered_data$rating - filtered_data$b_i - filtered_data$b_u -
                    filtered_data$b_y -  avg_rating_edx - filtered_data$b_genre

genre_bias_value <- sum(effects_value)/(nrow(filtered_data)+5)

edx_tidy <- edx_tidy %>% mutate(b_genre = 
                              ifelse(No_Genre == 1,genre_bias_value,0)+b_genre)

validation_tidy <-validation_tidy %>% mutate(b_genre =
                            ifelse(No_Genre == 1,genre_bias_value,0) + b_genre)

head(validation_tidy[,c(1:4,26:29)])


```


Pred_rating = avg_rating + b_i + b_u +b_y + b_genre

```{r}


pred_rating <- avg_rating_edx + validation_tidy$b_i + validation_tidy$b_u +
                                validation_tidy$b_y + validation_tidy$b_genre


```


# Result


Residual mean squared error RMSE is:

```{r}
RMSE(validation_tidy$rating,pred_rating)
```

# Conclusion

The RMSE is 0.8646 less than 0.8649 the minimum value put for the project and it means that we can trust our predicted values for movie ratings.







